---
title: "Correlation Matrix Analysis"
author: "Julian Mittmann-Goetsch"
date: "2024-08-06"
output: html_document
---

## Version 1.1: Updated version log format - 17.07.2025 | JM
## Version 1.0: Correlation matrix analysis for MERIT variables - 11.02.2025 | JM
## Purpose: Generate correlation matrices for EEA, community, and environmental variables

# Step 1: Prepare session and load data
## Step 1a: Install necessary packages
```{r, message=FALSE, warning=FALSE}
#install.packages("ggstatsplot")
#install.packages("tidyverse")
#install.packages("here")
```

## Step 1b: Load necessary packages
```{r}
library(ggstatsplot)
library(tidyverse)
library(here)
```

## Step 1c: Load processed datasets
```{r}
# Load EEA final dataset (created in script_eea.Rmd)
# This should contain the summarized EEA data with environmental variables
if(exists("df_EEA_final")) {
  df_eea_0_final <- df_EEA_final
  print("EEA final dataset loaded from environment")
} else {
  # Alternative: Load from saved file if available
  tryCatch({
    df_eea_0_final <- read.csv(here("data", "eea", "df_EEA_final.csv"))
    print("EEA final dataset loaded from file")
  }, error = function(e) {
    stop("EEA final dataset not found. Please run script_eea.Rmd first.")
  })
}

# Load other required datasets
# Community alpha diversity (from script_community.Rmd)
if(exists("df_alpha_sum")) {
  df_alpha_0_summary <- df_alpha_sum
  print("Alpha diversity dataset loaded from environment")
} else {
  tryCatch({
    df_alpha_0_summary <- read.csv(here("data", "rna", "df_alpha_summary.csv"))
    print("Alpha diversity dataset loaded from file")
  }, error = function(e) {
    warning("Alpha diversity dataset not found. Some correlations may be missing.")
    df_alpha_0_summary <- NULL
  })
}

# qPCR data (from script_qpcr_2023.Rmd)
if(exists("df_qpcr_sum")) {
  df_qpcr_0_summary <- df_qpcr_sum
  print("qPCR dataset loaded from environment")
} else {
  tryCatch({
    df_qpcr_0_summary <- read.csv(here("data", "qpcr", "df_qpcr_summary.csv"))
    print("qPCR dataset loaded from file")
  }, error = function(e) {
    warning("qPCR dataset not found. Some correlations may be missing.")
    df_qpcr_0_summary <- NULL
  })
}

# Warming data (from script_warming_data_2022.Rmd)
if(exists("df_warming_all_2022_monthly")) {
  df_warming_0_monthly <- df_warming_all_2022_monthly
  print("Warming dataset loaded from environment")
} else {
  tryCatch({
    df_warming_0_monthly <- read.csv(here("scripts", "df_warming_all_2022_monthly.csv"))
    print("Warming dataset loaded from file")
  }, error = function(e) {
    warning("Warming dataset not found. Some correlations may be missing.")
    df_warming_0_monthly <- NULL
  })
}

# IRIS data (from script_iris.Rmd)
if(exists("df_iris_summarized_2_final")) {
  df_iris_0_summary <- df_iris_summarized_2_final
  print("IRIS dataset loaded from environment")
} else {
  tryCatch({
    df_iris_0_summary <- read.csv(here("data", "iris", "df_iris_summary.csv"))
    print("IRIS dataset loaded from file")
  }, error = function(e) {
    warning("IRIS dataset not found. Some correlations may be missing.")
    df_iris_0_summary <- NULL
  })
}
```

# Step 2: Prepare correlation datasets
## Step 2a: Create comprehensive summary dataset
```{r}
# Start with EEA final data
df_corr_1_base <- df_eea_0_final %>%
  mutate(plotid = as.character(Plot))

# Add alpha diversity if available
if(!is.null(df_alpha_0_summary)) {
  df_alpha_0_summary <- df_alpha_0_summary %>% 
    mutate(plotid = as.character(plotid))
  
  df_corr_1_base <- df_corr_1_base %>%
    left_join(df_alpha_0_summary, by = c("plotid", "Layer"))
}

# Add qPCR data if available
if(!is.null(df_qpcr_0_summary)) {
  df_qpcr_0_summary <- df_qpcr_0_summary %>% 
    mutate(plotid = as.character(plotid))
  
  df_corr_1_base <- df_corr_1_base %>%
    left_join(df_qpcr_0_summary, by = "plotid")
}

# Add warming data if available
if(!is.null(df_warming_0_monthly)) {
  df_warming_0_monthly <- df_warming_0_monthly %>% 
    mutate(plotid = as.character(plotid))
  
  df_corr_1_base <- df_corr_1_base %>%
    left_join(df_warming_0_monthly, by = "plotid")
}

# Add IRIS data if available
if(!is.null(df_iris_0_summary)) {
  df_iris_0_summary <- df_iris_0_summary %>% 
    mutate(plotid = as.character(plotid))
  df_corr_1_base <- df_corr_1_base %>%
    left_join(df_iris_0_summary, by = c("plotid", "Layer"))
}

print("Base correlation dataset created")
print(paste("Dimensions:", nrow(df_corr_1_base), "rows x", ncol(df_corr_1_base), "columns"))
```

## Step 2b: Select and rename variables for correlation analysis
```{r}
# Select key variables and rename for better visualization
# Die Alpha-Diversity-Spalte heißt vermutlich nicht exakt 'Alpha_diversity'.
# Prüfe Alternativen wie 'alpha_diversity', 'AlphaDiversity', 'Shannon', etc.
# Passe ggf. den Spaltennamen hier an!
df_corr_2_selected <- df_corr_1_base %>%
  select(
    Layer,
    "Carbon acquisition" = Activity_C,
    "Nitrogen acquisition" = Activity_N,
    "Microbial abundance" = Abundance, # qPCR
    "Alpha diversity" = alpha_diversity, # ggf. anpassen!
    "Soil temperature" = pinTemp_mean,
    "Marsh surface elevation" = Elevation_mean,
    "Reduction index" = ri
  ) %>%
  # Remove rows with all NA values for numeric variables
  filter(!if_all(c("Carbon acquisition":"Reduction index"), is.na))

print("Variables selected for correlation analysis:")
print(colnames(df_corr_2_selected))
print(paste("Final dataset dimensions:", nrow(df_corr_2_selected), "rows x", ncol(df_corr_2_selected), "columns"))
```

# Step 3: Generate correlation matrices
## Step 3a: Overall correlation matrix (all layers combined)
```{r}
# Remove Layer column for overall analysis
df_corr_overall <- df_corr_2_selected %>%
  select(-Layer)

# Generate overall correlation matrix
plot_corr_overall <- ggstatsplot::ggcorrmat(
  data = df_corr_overall,
  matrix.type = "lower",
  type = "parametric", # Use parametric for Pearson correlation
  colors = c("darkred", "white", "steelblue"),
  title = "Overall Correlation Matrix - All Variables",
  subtitle = "Pearson correlation coefficients with significance levels"
)

print("Overall correlation matrix generated")
plot_corr_overall
```

## Step 3b: Layer-specific correlation matrices
```{r}
# Generate grouped correlation matrix by layer
plot_corr_by_layer <- ggstatsplot::grouped_ggcorrmat(
  data = df_corr_2_selected,
  grouping.var = Layer,
  matrix.type = "lower",
  type = "parametric", # Use parametric for Pearson correlation
  colors = c("darkred", "white", "steelblue"),
  title.prefix = "Correlation Matrix",
  subtitle.prefix = "Pearson correlations"
)

print("Layer-specific correlation matrices generated")
plot_corr_by_layer
```

# References
```{r}
# ggstatsplot: Patil, I. (2021). Visualizations with statistical details: The 'ggstatsplot' approach. Journal of Open Source Software, 6(61), 3167, doi:10.21105/joss.03167
# tidyverse: Wickham, H., Bryan, J., François, R., Henry, L., Müller, K., Pedersen, T. L., & Takahashi, K. (2024). tidyverse: Easily Install and Load the 'Tidyverse'. R package version 2.0.0. https://CRAN.R-project.org/package=tidyverse
# here: Müller, K. & Bryan, J. (2024). here: A Simpler Way to Find Your Files. R package version 1.0.1. https://CRAN.R-project.org/package=here
```

---

# REVISION - September 2025

## Step 4: Comprehensive summary table by Zone, Treatment, and Layer
```{r}
df_corr_3_summary_table <- df_corr_1_base %>%
  mutate(
    Zone = factor(Zone, levels = c("Pioneer zone", "Low marsh", "High marsh")),
    Treatment = factor(Treatment, levels = c("ambient", "+1.5°C", "+3.0°C")),
 
  ) %>%
  group_by(Zone, Treatment, Layer) %>%
  summarise(
    across(where(is.numeric), list(mean = ~mean(.x, na.rm = TRUE), 
                                   se = ~sd(.x, na.rm = TRUE)/sqrt(sum(!is.na(.x))))),
    .groups = "drop"
  )

write.csv(df_corr_3_summary_table, here("tables", "comprehensive_summary_all_variables.csv"), row.names = FALSE)
```

---

## REVISION DOCUMENTATION - September 2025

### Changes Made During Revision:
1. **Added Step 4: Comprehensive Summary Table**
   - Created detailed summary statistics grouped by Zone, Treatment, and Layer
   - Calculated means and standard errors for all available variables
   - Implemented flexible variable detection to handle missing datasets

2. **Variables Included:**
   - EEA variables: Carbon acquisition, Nitrogen acquisition, C:N ratio
   - Microbial metrics: Bacterial abundance (qPCR), Alpha diversity (Shannon)
   - Environmental variables: Soil temperature, Elevation, Reduction index

3. **Output Files Generated:**
   - `tables/comprehensive_summary_all_variables.csv`: Complete numeric summary
   - `tables/comprehensive_summary_formatted.csv`: Formatted with mean ± SE notation

4. **Statistical Framework:**
   - Grouped by experimental design factors (Zone × Treatment × Layer)
   - Standard error calculations for all variables
   - Proper factor level ordering for consistent presentation

**Rationale**: Provides comprehensive overview of all measured variables across the experimental design for manuscript tables and supplementary materials.

*Last modified: September 15, 2025 | JM*
