---
title: "qPCR Analysis - Bacterial Abundance (16S rRNA)"
author: "Julian Mittmann-Goetsch"
date: "2025-01-31"
output: html_document
---

## Version 2.1: Updated version log format - 17.07.2025 | JM
## Version 2.0: Unified script structure, English comments, hierarchical naming - 07.07.2025 | JM
## Version 1.1: Streamlined script - 02.06.2025 | JM


# Step 1: Prepare session and load data
## Step 1a: Load necessary packages
```{r}
library(emmeans)
library(ggplot2)
library(ggpubr)
library(here)
library(lme4)
library(performance)
library(tidyverse)
```

## Step 1b: Load qPCR data
```{r}
# Load bacterial abundance data from qPCR analysis
df_eub_0_raw <- read.delim(here("data", "qpcr", "df_copies_EUB.txt"), dec=".", sep="\t")
```

## Step 1c: Define plot theme
```{r message=FALSE, warning=FALSE}
# Global theme for all plots in this analysis
theme_JM <- theme(
  axis.title.x      = element_text(colour = "black", size = 12),
  axis.text.x       = element_text(colour = "black", size = 12), 
  axis.title.y      = element_text(colour = "black", size = 12),
  axis.text.y       = element_text(colour = "black", size = 12), 
  legend.title      = element_text(colour = "black", size = 12),
  legend.text       = element_text(colour = "black", size = 12), 
  legend.position   = "top",
  panel.background  = element_blank(), 
  panel.border      = element_rect(colour = "black", fill = NA, size = 1.2),
  legend.key        = element_blank(),
  strip.text.x      = element_text(colour = "black", size = 12),
  strip.text.y      = element_text(colour = "black", size = 12)
)
```

# Step 2: Data preprocessing
## Step 2a: Process and format data
```{r}
# Apply hierarchical naming and factor processing
df_eub_1_processed <- df_eub_0_raw %>%
  mutate(
    Depth = recode_factor(Depth, "0-10" = "0-10 cm", "20-30" = "20-30 cm"),
    Treatment = recode_factor(Treatment, "ambient" = "ambient", "1.5C" = "+1.5°C", "3.0C" = "+3.0°C"),
    Zone = dplyr::recode(Zone, "PZ" = "Pioneer zone", "LM" = "Low marsh", "HM" = "High marsh"),
    Zone = factor(Zone, levels = c("Pioneer zone", "Low marsh", "High marsh")),
    core_id = paste0(Plot, Subplot)
  ) %>%
  filter(Subplot == "ctr")  # Remove root-barrier samples
```

# Step 3: Data visualization
## Step 3a: Create barplot for bacterial abundances
```{r}
# Generate barplot showing bacterial abundances by zone and treatment
plot_qpcr_bar_eub <- ggplot(df_eub_1_processed, aes(x = Zone, y = log10_copies, fill = Treatment)) +
  scale_y_continuous(limits = c(9, 10)) + 
  stat_summary(fun = mean, geom = "bar", position = position_dodge(width = 0.9), alpha = 0.9, width = 0.7) +
  stat_summary(fun.data = mean_se, geom = "errorbar", position = position_dodge(width = 0.9), width = 0.25, color = "black") +
  scale_fill_manual(values = c("ambient" = "#4575B4", "+1.5°C" = "#FDAE61", "+3.0°C" = "#D73027"), name = "Warming treatment:") +
  labs(fill = "Treatment", y = "log10 abundance 16S rRNA [copies/g OM]", x = "Zone") +
  theme_JM 
# Display plot
print(plot_qpcr_bar_eub)
```

# Step 4: Statistical analysis
## Step 4a: Linear mixed-effects model for bacterial abundance
```{r}
# Fit linear mixed-effects model with treatment and zone interaction
stat_model_eub <- lmer(log10_copies ~ Treatment * Zone + (1|core_id), data = df_eub_1_processed)

# Model diagnostics
cat("Model diagnostics:\n")
print(performance::check_model(stat_model_eub))

# ANOVA results
cat("\nANOVA results:\n")
print(anova(stat_model_eub))

# Model summary
cat("\nModel summary:\n")
print(summary(stat_model_eub))
```

## Step 4b: Post-hoc comparisons
```{r}
# Pairwise comparisons between zones
stat_emm_zones <- emmeans(stat_model_eub, ~ Zone)
stat_pairs_zones <- pairs(stat_emm_zones)

cat("Zone comparisons:\n")
print(stat_pairs_zones)

# Pairwise comparisons between treatments within zones
stat_emm_treatment_zone <- emmeans(stat_model_eub, ~ Treatment | Zone)
stat_pairs_treatment_zone <- pairs(stat_emm_treatment_zone)

cat("\nTreatment comparisons within zones:\n")
print(stat_pairs_treatment_zone)
```

# Step 5: Data summary
## Step 5a: Create summary statistics
```{r}
# Summary by zone and treatment
df_eub_2_summary <- df_eub_1_processed %>%
  group_by(Zone, Treatment) %>%
  summarise(
    mean_abundance = mean(log10_copies, na.rm = TRUE),
    se_abundance = sd(log10_copies, na.rm = TRUE)/sqrt(n()),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(mean_se = paste0(round(mean_abundance, 2), " ± ", round(se_abundance, 2), " (n=", n, ")"))

cat("Summary statistics by zone and treatment:\n")
print(df_eub_2_summary)

# Summary by plot for export
df_eub_3_plot_summary <- df_eub_1_processed %>%
  mutate(plotid = as.factor(Plot)) %>%
  group_by(plotid) %>%
  summarise(
    Abundance = mean(log10_copies, na.rm = TRUE),
    .groups = "drop"
  )

cat("\nSummary statistics by plot:\n")
print(df_eub_3_plot_summary)

# Export summarized qPCR data for correlation matrix integration
write.csv(df_eub_3_plot_summary, here("data", "qpcr", "df_qpcr_summary.csv"), row.names = FALSE)
```

# References
```{r}
# emmeans: Lenth, R. V. (2024). emmeans: Estimated Marginal Means, aka Least-Squares Means. R package version 1.10.2. https://CRAN.R-project.org/package=emmeans
# ggplot2: Wickham, H., Chang, W., Henry, L., Pedersen, T. L., Takahashi, K., Wilke, C., Woo, K., & Yutani, H. (2024). ggplot2: Create Elegant Data Visualisations Using the Grammar of Graphics. R package version 3.5.1. https://CRAN.R-project.org/package=ggplot2
# ggpubr: Kassambara, A. (2024). ggpubr: 'ggplot2' Based Publication Ready Plots. R package version 0.6.0. https://CRAN.R-project.org/package=ggpubr
# here: Müller, K. & Bryan, J. (2024). here: A Simpler Way to Find Your Files. R package version 1.0.1. https://CRAN.R-project.org/package=here
# lme4: Bates, D., Mächler, M., Bolker, B., & Walker, S. (2024). lme4: Linear Mixed-Effects Models using 'Eigen' and S4. R package version 1.1-35. https://CRAN.R-project.org/package=lme4
# performance: Lüdecke, D., Ben-Shachar, M. S., Patil, I., Waggoner, P., & Makowski, D. (2024). performance: Assessment of Regression Models Performance. R package version 0.10.8. https://CRAN.R-project.org/package=performance
# tidyverse: Wickham, H., Bryan, J., François, R., Henry, L., Müller, K., Pedersen, T. L., & Takahashi, K. (2024). tidyverse: Easily Install and Load the 'Tidyverse'. R package version 2.0.0. https://CRAN.R-project.org/package=tidyverse
```

